Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-09-17 closed_monoidal/closed_monoidal.tex
updated slides and closed monoidal notes

\begin{example}
  Let $\mathcal{C} = (\mathcal{C}_{0},I,\otimes, l, r, \alpha)$ be a monoidal category and consider the bifunctor $\otimes : \mathcal{C} \times \mathcal{C} \to \mathcal{C}$ on $\mathcal{C}$.
  It induces an enriched monoidal category $\enriched{C} = (\enriched{C}_{0}, I, \enriched{\otimes}, \overline{l}, \overline{r}, \enriched{\alpha})$ such that
  \begin{enumerate}
    \item $A \overline{\otimes} B \coloneq A \otimes B$;
    \item $\overline{l} : A \overline{\otimes} I \to A \coloneq I \to \enriched{C}(A \overline{\otimes} I, I)$, i.e. $\overline{l}$ picks out $l$ out of $\enriched{C}(A \overline{\otimes} I, I)$;
    \item similarly for $\overline{r}$, $\overline{\alpha}$.
  \end{enumerate}
  $\otimes_{A,B,C,D} : \mathcal{C}(A,B) \times \mathcal{C}(C,D) \to \mathcal{C}(A \otimes C, B \otimes D)$ becomes $\overline{\otimes}_{A,B,C,D} : \enriched{C}(A,B) \times \enriched{C}(C,D) \to \enriched{C}(A \overline{\otimes} C, B \overline{\otimes} D)$.
  In particular, $f + g \;\overline{\otimes}\; h + k = (f \;\overline{\otimes}\; h + k) + (g \;\overline{\otimes}\; h + k) = (f \;\overline{\otimes}\; k) + (f \;\overline{\otimes}\; h) + (g \;\overline{\otimes}\; h) + (g \;\overline{\otimes}\; k)$, where $+$ binds tighter than $\overline{\otimes}$.
\end{example}

@@

By freely enriching $\mathcal{C}$ we obtain an enriched trace
\[
\text{tr}^{X} : \enriched{C}(A \overline{\otimes} X, B \overline{\otimes} X) \to \enriched{C}(A,B)
\]
and all the equations are preserved with the addition of a new equation
\[
  \text{tr}^{X}(f + g) = \text{tr}^{X}(f) + \text{tr}^{X}(g)
\]
for $f,g : A \overline{\otimes} X \to B \overline{\otimes} X$
TODO: trace?

>>>>

\begin{example}
  Let $\mathcal{C} = (\mathcal{C}_{0},I,\otimes, l, r, \alpha)$ be a monoidal category and consider the bifunctor $\otimes : \mathcal{C} \times \mathcal{C} \to \mathcal{C}$ on $\mathcal{C}$.
  It induces an enriched monoidal category $\enriched{C} = (\enriched{C}_{0}, I, \enriched{\otimes}, \overline{l}, \overline{r}, \enriched{\alpha})$ such that
  \begin{enumerate}
    \item $A \overline{\otimes} B \coloneq A \otimes B$;
    \item $\overline{l} : A \overline{\otimes} I \to A \coloneq I \to \enriched{C}(A \overline{\otimes} I, I)$, i.e. $\overline{l}$ picks out $l$ out of $\enriched{C}(A \overline{\otimes} I, I)$;
    \item similarly for $\overline{r}$, $\overline{\alpha}$.
  \end{enumerate}
  $\otimes_{A,B,C,D} : \mathcal{C}(A,B) \times \mathcal{C}(C,D) \to \mathcal{C}(A \otimes C, B \otimes D)$ becomes $\overline{\otimes}_{A,B,C,D} : \enriched{C}(A,B) \times \enriched{C}(C,D) \to \enriched{C}(A \overline{\otimes} C, B \overline{\otimes} D)$.
  In particular, $f + g \;\overline{\otimes}\; h + k = (f \;\overline{\otimes}\; h + k) + (g \;\overline{\otimes}\; h + k) = (f \;\overline{\otimes}\; k) + (f \;\overline{\otimes}\; h) + (g \;\overline{\otimes}\; h) + (g \;\overline{\otimes}\; k)$, where $+$ binds tighter than $\overline{\otimes}$.
  \question{I need to show that $\overline{\otimes}$ is a bifunctor. Functoriality in each argument does not entail bifunctoriality. This is the difference between monoidal and pre-monoidal categories.
  }
\end{example}

@@

By freely enriching $\mathcal{C}$ we obtain an enriched trace
\[
\text{tr}^{X} : \enriched{C}(A \overline{\otimes} X, B \overline{\otimes} X) \to \enriched{C}(A,B)
\]
and all the equations are preserved with the addition of a new equation
\[
  \text{tr}^{X}(f + g) = \text{tr}^{X}(f) + \text{tr}^{X}(g)
\]
for $f,g : A \overline{\otimes} X \to B \overline{\otimes} X$

\section{E-Hypergraphs for $\catname{SLat}$-enriched symmetric closed monoidal categories}

We can extend the definition of e-hypergraphs for $\catname{SLat}$-enriched symmetric monoidal categories~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} to accommodate new structure brought by closedness.

\begin{definition}
Closed monoidal signature $\Sigma$ is a pair $(\Sigma_{O},\Sigma_{M})$ of signatures for objects and signatures for morphisms with sources and targets in $\text{obj}_{\Sigma_{O}}$ defined as
\begin{itemize}
  \item $I$ (monoidal unit) is in $\text{obj}_{\Sigma_{O}}$
  \item For every $o \in \Sigma_{O}$, $o \in \text{obj}_{\Sigma_{O}}$;
  \item For every $o_1$ and $o_2$ $ \in \text{obj}_{\Sigma_{O}}$, $o_1 \otimes o_2 \in \text{obj}{\Sigma_{O}}$ and $o_1 \multimap o_2 \in \text{obj}_{\Sigma_{O}}$
\end{itemize}
\end{definition}

\begin{definition}
 Closed symmetric monoidal theory ($\catname{CSMT}$) over a closed monoidal signature $\Sigma$ and equations $\mathcal{E}$ denoted as $\catname{CSMT}(\Sigma, \mathcal{E})$ is given by a set of terms $T$
 \begin{align*}
 T \coloneq\; &t : A \to B \in \Sigma_{M} \;|\; t_1 \otimes t_2 : A \otimes C \to B \otimes D \text{ for } t_{1} : A \to B, t_{2} : C \to D \in T\\
            & \;|\; t_{1};t_{2} : A \to C \text{ for } t_{1} : A \to B, t_{2} : B \to C \in T \; | \;  \Lambda(t) : A \to X \multimap B \text{ for } t : A \otimes X \to B \\
            &  \;|\; \text{eval}(t_1 \otimes t_2) : B \text{ for } t_1 : X \multimap B, t_2 : X \in T \\
            &  \;|\; \text{sym}_{A,B} \;|\; id_{A}
 \end{align*}
 quotiented by the laws of a closed symmetric monoidal category.
\end{definition}
\question{Maybe types and terms combined is not a good idea}
\begin{definition}

We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{red}{<}, \textcolor{blue}{<},\consistency)$ where
\begin{itemize}
  \item $V$ is a set of vertices;
  \item $E = \textcolor{gray}{E} \cup \textcolor{red}{E} \cup \textcolor{blue}{E}$ is a set of hyperedges that include \textit{plain} edges and two types of \textit{hierarchical} edges respectively;
  \item $s,t : E \to V^{*}$ are source and target functions;
  \item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \text{obj}_{\Sigma_{O}}$ are \textit{label} functions;
  \item $\textcolor{red}{<} : V + E \to \textcolor{red}{E}$ and $\textcolor{blue}{<} : V + E \to \textcolor{blue}{E}$ are partial \textit{child} functions that are disjoint: if $\textcolor{red}{<}(x)$ is defined then $\textcolor{blue}{<}(x)$ is undefined and vice-versa;
  \item $\consistency = \bigcup_{p} \consistency_{p}$, where $\consistency_{p}$ is a \textit{consistency} relation defined on each set $\{x \in V + E \;|\; p \;\textcolor{red}{<^{\mu}}\; x\}$;
\end{itemize} 
\end{definition}

$\textcolor{red}{E}$ will contain edges that encode the equivalence classes and $\textcolor{blue}{E}$ the ones that defined the lambda abstraction.
All the above functions and relations must satisfy the obvious properties (TODO: add them).

Similarly, we have a category of (extended) cospans formed by such e-hypergraphs where morphisms are of the form
\[
n \xrightarrow{f_{\text{ext}}} n' \xrightarrow{f_{\text{int}}} \mathcal{G} \xleftarrow{g'_{\text{int}}} m' \xleftarrow{g_{\text{ext}}} m
\]

where $n$ and $m$ are external input and output interfaces respectively and $n'$, $m'$ are input and output internal interfaces.
$n,n',m,m'$ are discrete ordered e-hypergraphs.
We further require that $f_{\text{ext}}$ and $g_{\text{ext}}$ are monos and all vertices in the image of $f_{\text{ext}};f_{\text{int}}$ (respectively, $g_{\text{ext}};g_{\text{int}}$) are top-level.

To build a correspondence between terms of a closed SMC and e-hypergraphs with interfaces (above) we restrict the cospans of the latter to monogamous cospans.

\begin{definition}
  We call a cospan 
  \[
n \xrightarrow{f_{\text{ext}}} n' \xrightarrow{f_{\text{int}}} \mathcal{G} \xleftarrow{g'_{\text{int}}} m' \xleftarrow{g_{\text{ext}}} m
\]
\textit{weak} monogamous if
\begin{itemize}
  \item in- and out- degrees of every vertex is at most 1;
  \item $f_{\text{int}}$ and $g_{\text{int}}$ are monos;
  \item Vertices with in-degree (respectively, out-degree) of 0 are precisely the image of $f_{\text{int}}$ (respectively, $g_{\text{int}}$).
\end{itemize}

\end{definition}

\begin{definition}[Isomorphic cospans]
Consider a relation 
\[
R = \{ x R y \text{ if } f_{\text{int}}(x) \consistency f_{\text{int}}(y) \;\text{or}\; \exists z \;\text{s.t.}\; z \textcolor{blue}{<^{\mu}} f_{\text{int}}(x) \text{ and } z \textcolor{blue}{<}^{\mu} f_{\text{int}}(y)\}
\]
for $x, y \in n'$.
And let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}^{k}_{j=1}$.
We get an analogous partition for $m'$.

Two extended cospans are isomorphic if there exist isomorphisms $\alpha$, $\beta$ and $\gamma$ making the following diagram commute and such that $\alpha$ and $\gamma$ preserve order within each $p_j$.
\[
\scalebox{0.8}{
    \tikzfig{../figures/combinatorial_semantics/isomorphic_e_cospans}
}
\]
\end{definition}

Consider an example of isomorphic and non-isomorphic cospans below and their respective string diagrams

\begin{figure}
  \begin{subfigure}[c]{0.4\linewidth}
    \scalebox{0.6}{
    \tikzfig{./figures/closed_iso_example_1}
    }
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{0.4\linewidth}
    \scalebox{0.6}{
    \tikzfig{./figures/closed_iso_example_2}
    }
  \end{subfigure}
\end{figure}

\begin{definition}

We define ordered sets of \textit{input} (respectively, \textit{output}) vertices of a hierarchical edge $e$ as the vertices $v_{i}$ such that $e < v_{i}$ which are in the image of $f_{\text{int}}$ (respectively, $g_{\text{ext}}$) with the induced ordering.
\end{definition}

\begin{definition}

Given an ordered set $S$ of labelled vertices we can form words formed by concatenating the labels.
We will denote such words as $w(S)$.
\end{definition}

\begin{definition}
We call a monogamous cospan
\[
  n \xrightarrow{f_{\text{ext}}} n' \xrightarrow{f_{\text{int}}} \mathcal{G} \xleftarrow{g'_{\text{int}}} m' \xleftarrow{g_{\text{ext}}} m
\]
\textit{well-typed} if all hierarchical edges of $\mathcal{G}$ are well-typed in the sense below.
\begin{itemize}
  \item For each $e \in \textcolor{red}{E}$ consider sets $I$ and $O$ of its input and output vertices partitioned according to $\consistency_{e}$.
        $e$ is well-typed if for each element $S$ of the partition of $I$ $w(S) = w(s(e))$ and similarly for $O$ and $t(e)$.
  \item For each $e \in \textcolor{blue}{E}$ consider sets $I$ and $O$ of its input and output vertices.
        $e$ is well-typed if there exists an object $B \in \text{obj}_{\Sigma_{O}}$ such that $[...s(e), B] = w(I)$ and $B \multimap w(O) = w(t(e))$
\end{itemize}
\end{definition}

### 2024-09-17 slides/slides.tex
slides update
updated slides and closed monoidal notes

\[
\Sigma = \{id_n : n \to n,\sigma_{n,m} : n + m \to m + n,\mu : 1 \to 2, i : 1 \to 0\}
\]

@@

\begin{minted}{c++}
    enum CommutativeComonoids {
        "mu" = Mu([Id;1]),
        "i" = I([Id;1]),
        ";" = Compose([Id;2]),
        "parallel" = Parallel([Id;2]),
        "id" = Identity([Id;1]),
        "I" = Unit,
        "sym" = Sym([Id;2])   
    }
\end{minted}

@@

        Before saturation
        \vspace{-3em}
        \begin{figure}
            \includegraphics[scale=0.4]{figures/egraph_before_saturation.pdf}
        \end{figure}
    \end{example}
\end{frame}

>>>>

\[
\Sigma = \{id_0 : 0 \to 0, id_{1} : 1 \to 1, \sigma_{1,1} : 1 + 1 \to 1 + 1,\mu : 1 \to 2, i : 1 \to 0\}
\]

@@

\begin{minted}{c++}
    enum CommutativeComonoids {
        "mu" = Mu, // mu : 1 -> 2
        "i" = I, // i : 1 -> 0
        ";" = Compose([Id;2]),
        "parallel" = Parallel([Id;2]),
        "id" = Identity1,
        "id_2" = Identity2,
        "I" = Unit,
        "sym_1_1" = Sym1_1,
        "sym_2_2" = Sym2_2,
        "sym_2_1" = Sym2_1,
        "sym_1_2" = Sym1_2,
        Symbol(egg::Symbol),
    }
\end{minted}

@@

        Before saturation
        \vspace{-3em}
        \begin{figure}
            \includegraphics[scale=0.4]{figures/egraph_before_saturation_fix.pdf}
        \end{figure}
    \end{example}
\end{frame}
